\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-04-05 paper.tex
Update introduction

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
methods are all based, in different ways, on inferring
approximate \emph{structures}
instead of a complete and fully detailed history
(Appendix~\ref{sec-survey-arg-infer}).
It is important to distinguish here between
structures and models: whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
One could heuristically estimate a fully precise ARG,
% this would be a cool thing to do!
or statistically sample a partial,
approximate ARG under a model
such as the coalescent.
Although the term ``ARG'' is now often used in a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied approximate structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank},
there is no corresponding mathematical definition
that is sufficiently general.

(see \citet{lewanski2024era,brandt2024promise} for biologically oriented
introductions)

We highlight this problem and show how to address it by a simple formal definition of an
ARG data structure, based on recording the intervals of
genetic inheritance between specific genomes.
We call this the ``genome ARG'', or gARG encoding. We
contrast this with the classical formal definition of an ARG, based on
recording common ancestor and recombination events, which we refer to as the
``event ARG'' or eARG encoding. We show that the new gARG encoding is a
substantial generalisation of the classical eARG approach,  providing much more
flexibility in how genetic inheritance can be represented, and encompasses the
outputs of modern methods. We show that the gARG approach can represent many
different types of approximation, in particular allowing us to systematically
describe uncertainty about the temporal ordering of multiple recombinations.

>>>>

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is some
confusion about what, precisely, an ARG \emph{is}.
% The grammar is a nightmare here - the stochastic process is singular, right?
In its original form,
developed by Griffiths and colleagues,
the ARG is an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised
as a graph~\citep{griffiths1991two,
ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
% Move into plural, an ARG is an instance of the data structure in this view
Subsequently, an ARG has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
a \emph{realisation} of such a random process,
or an inferred ancestry of a sample of genomes.
The distinction between stochastic process
and data structure is not clear cut, however, and subfields use the term
differently (Appendix~\ref{sec-arg-history},\ref{sec-big-and-little-arg}).
% TODO rework these two sentences. Point about models vs structures 
% is an important one, we need to make that.
Recent large-scale methods infer approximate \emph{structures}
instead of a complete and fully detailed history
(Appendix~\ref{sec-survey-arg-infer}), and it is
therefore important to distinguish between
approximate structures and approximate models.
Whether an inference method is based
on heuristics or a rigorous mathematical model is orthogonal
to the level of detail provided in its estimate.
The term ``ARG'' therefore has many different meanings, varying 
over time and depending on context. 
There is an emerging consensus to use the term
in quite a general sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
informally
encompassing the varied structures output by modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient,
zhang2023biobank}. 
However, there is currently no formal definition or systematic 
discussion that unifies these different structures, 
ultimately stifling progress in this vibrant research area.

In this perspective we provide a simple formal definition
of an ARG data structure, which generalises classical 
definitions and encompasses the output of modern simulation and inference 
methods. We show that different levels of approximation
are possible using this structure, illustrated via examples.
% This is a bit crap, but do want to make this point about tskit for the 
% "only read the introduction" crowd.
The proposed ARG definition is the basis of the widely-used \texttt{tskit}
library which provides a powerful software platform
for ARG-based analysis and, we argue, would be a useful community standard.
This perspective is intended for ``ARG practitioners'', who we hope
will find the detailed examples and technical appendices useful.
Readers seeking an introduction to ARGs and their applications are 
directed to \citet{lewanski2024era}
and \citet{brandt2024promise}.
